Weak interspecific interactions in a sagebrush steppe: evidence from observations, models, and experiments

### 2017-09-10 removals/manuscript/ResponseToReviews.tex

Introduction

Methods

Lines 137-139:  When did you survey the removal plots? What month and for how many years?

\response{We added the following sentence, "We repeated mapped censuses of all plots each June from 2012-2016."}

@@

Lines 150-173: In this section, I was confused what plots (survey plots or removal experiment) you were using to fit the models. Originally I had thought you would use only the survey plots to fit the models, then hold those parameters constant and add the removal parameter and fit the models to the removal data. I had thought that you would want to separate the data used to parameterize the model vs. the data used to test the model. I then realized that you were fitting the model to all data simultaneously (survey and removal experiment). Maybe be explicit somewhere?

Were the 2011 pretreatment data in the removal plots used as the conditions for t=1 in the model (i.e. for the competitive effects)? I don’t think it should be, because for the majority of the 2011-2012 transition, the target species were growing without competition from the removed species. Thus you might find that when there was a lot of Artemisia pretreatment in 2011, there is a large positive growth response in 2012.

\response{We added this sentence to the Methods: "For plants in removal plots in 2012, we set crowding for removed neighbor species to zero, rather than using the pre-treatment (2011) data."}

@@

\response{Thank you, fixed.}

Tables and Figures:

Fig. 1: I’m confused how the focal species account for 70\% basal and 60\% canopy cover (as stated in the methods), but the mean cover values for the grasses in Fig. 1 range from only 1-3\%. I realize your estimates of cover were highly precise, but it seems that it would be hard to model species that are at such low abundance.

\response{We addressed this point in detail in response to Reviewer 2's comment below.}

@@

The test of the effects of shrubs on grasses is stronger (mean initial shrub cover was ca. 30\% before
removal) and indeed two out of three tests found significant shrub removal effects on growth beyond
the regressions on neighborhood crowding, as well as on population growth and the removal model
provides better predictions from the IBM for P. spicata. However, the authors argued this result may
not reflect stronger interspecific competition than estimated by the model because the release was no
greater for plants originally under a canopy than outside a sagebrush canopy. However, that argument
only makes sense either if competition is only for light or if the canopy edge reflects the edge of the
rooting zone as well. Given the habitat, I doubt the former is true and the authors also suggest it is
unlikely (line 442). For the latter, is anything known about the relationship of root distribution to
canopy cover in A. tripartita? My guess is that roots extend well beyond the canopy edge in these
habitats, leaving underestimated competitive interactions as a likely explanation for the significant
removal effect.

\response{This is a fair point. We added a sentence noting that grasses in interspaces could still experience competition
if shrub roots extend beyond canopies.}

@@

Given this result for P. spicata, line 401 of the discussion is a bit surprising: “how to reconcile the
insensitivity of grasses to sagebrush removal…..” On the other hand, it does hold for two of the grasses
and some discussion of why the grass species differ in sensitivity to shrub removal would be interesting.
Regardless of which species it holds for, the authors then attempt to reconcile the insensitivity found in
this study with strong effects found in the rangeland ecology literature by noting the relatively low shrub
cover (and small size of A. tripartita) in these ungrazed areas to explain the lack of strong effect of shrub
removal on grasses (lines 401-412). I agree this is a reasonable explanation and would argue that this is
also a possible, even likely, explanation for lack of effect of the grasses on sagebrush as above. In either
case, this reason for weak interspecific interactions (few interspecific neighbors) is quite different from
the ecological explanations in the next paragraph (lines 413-426; e.g., resource partitioning, balance of
facilitation and competition) and has quite different implications from those discussed in the
subsequent paragraph (lines 427-438). If this material stays, it would be helpful to compare and
contrast these different sorts of explanations and their implications more directly.

\response{We thank the reviewer for pointing out this apparent contradiction. In fact, consistent with 
the reviewers' critique of our argument, we actually demonstrated that 
artificially increasing Artemisia cover would not lead to a much bigger competitive release. We have added text to this 
section to suggest that overgrazing must do more than simply increase shrub cover, but must also alter
the per capita competition coefficients, potentially by altering the balance between competition and facilitation
or changing resource availability. }

### 2017-09-10 removals/manuscript/removal_ms.tex
Still working on revisions, just saving to be safe

In spring 2011, in a large exclosure containing six of the historical permanent quadrats, we established an additional 16 quadrats for the removal treatments. New quadrats were selected to be similar to historical plots, but also needed to have enough of the focal species present so that removal experiments could be conducted. We evaluated potential locations for the new quadrats near the original quadrats and then used only those locations with at least 5\% canopy cover of \textit{Artemisia tripartita} and at least 3\% combined basal cover of the three focal perennial grasses. We rejected locations falling on hill slopes, areas with greater than 20\% bare rock, or with over 10\% cover of the woody shrubs \textit{Purshia tridentata} or \textit{Amelanchier utahensis}. We mapped the new quadrats in June, 2011 and then implemented the removal treatments in September, 2011. We randomly assigned 8 of the sixteen new quadrats to the sagebrush removal treatment, and the remaining 8 to the grass removal treatment. We cut sagebrush stems at ground level and applied a 5\% solution of Roundup Pro (Glyphosate, Monsanto) to the cut stems. In the grass removal plots, we painted the same herbicide solution on all perennial grasses. We removed plants 50 cm beyond the quadrat boundaries to minimize edge effects. The sagebrush removal was virtually 100\% effective; grass removals required additional herbicide applications in April and May of each year, with fewer grasses remaining each year.

@@

Our regression models for survival, growth and recruitment follow previous work \citep{adler_coexistence_2010,chu_large_2015}. The current analysis departs from previous work with the inclusion of a categorical removal treatment effect. Although the three vital rate models differ in functional form (see Appendix A for complete model descriptions), all of them have a linear predictor, $\mu$, containing the following terms:
\begin{equation}
\mu_{ij} = \gamma_j+ \varphi_{jg}+  \chi_j  + \sum \limits_{m} \omega_{jm} {W}_{ijm}
\label{eqn:survReg}
\end{equation}
where $i$ refers to an individual genet (for survival and growth) or quadrat (for recruitment) of species $j$, $\gamma$ is a time-dependent intercept (year effect), $\varphi$ is the coefficient for the effect of spatial location (`$g$' represents a group of nearby quadrats located within one pasture or grazing exclosure), $\chi$ is the removal treatment effect, 
$W_{ijm,t}$ is the crowding on species $j$ exerted by neighbors of species $m$ at the individual (survival and growth) or
quadrat (recruitment) scale, and $\omega_{jm}$ is an interaction coefficient which determines the impact of crowding by species $m$ on the 
focal species $j$.

@@

The story is different for the fourth species, \textit{P. spicata}. Both individual growth and projected short- and long-term population 
growth were higher in the sagebrush removal treatments than in the control plots, suggesting that our analyses of observational data may 
underestimate the competitive effect of \textit{A. tripartita} on \textit{P. spicata}, and the resulting positive effect of \textit{A. tripartita} 
removal. However, the mechanism by which \textit{A. tripartita} removal promotes \textit{P. spicata} remains unclear. If  \textit{A. tripartita} 
removal is releasing \textit{P. spicata} from competition, we should expect \textit{P. spicata} individuals that were located beneath \textit{A. tripartita} 
shrubs that we removed to respond more than individuals located away from any shrubs. Yet we found no evidence for such a difference. 
An alternative explanation involves pre-treatment differences in composition between control plots and  the \textit{A. tripartita} 
removal plots, which had higher cover of \textit{A. tripartita} (see 2011 means in Fig. \ref{fig:CoverTrends}) and higher cover of subdominant 
grasses and shrubs (Fig. \ref{fig:W-by-treatment}). Perhaps these pre-treatment differences reflect subtle differences in 
edaphic conditions which could promote \textit{P. spicata} growth. In summary, while \textit{P. spicata} appears to 
show stronger competitive release than our baseline model predicted, the evidence is not conclusive.

How can we reconcile the insensitivity of grasses to sagebrush removal in our study with evidence from rangeland ecology 
research showing that sagebrush eradication increases grass production? The most likely explanation involves ecological context: our 
study focused on \textit{Artemisia tripartita} in an ungrazed plant communities, whereas most research on sagebrush 
and forage production has focused on \textit{Artemisia tridentata} in heavily grazed communities, and often has involved grass seeding 
following sagebrush removal (e.g. \citealt{Cook1963}). \textit{A. tripartita} is smaller in stature than \textit{A. tridentata} 
and may interact differently with perennial grasses. More importantly, cover of sagebrush can increase dramatically on overgrazed sites, 
especially under a spring grazing regime. Competitive release of grasses following sagebrush removal often increases in 
a nonlinear fashion, with little increase in production at sites with low initial sagebrush cover, 
and accelerating responses as initial sagebrush cover passes some threshold \citep{McDaniel2005}. The cover of 
sagebrush in our ungrazed quadrats may have fallen below that threshold.

The weak interspecific interactions among many of the dominant species in our system implies that these species  
are limited by quite different factors. This low niche overlap could reflect strong resource partitioning \citep{tilman_resource_1982}, 
perhaps combined with variation in the times and locations at which species are taking up those resources. 
Species-specific responses to spatial and temporal variation in non-resource environmental variables can also weaken interspecific 
interactions \citep{chesson_mechanisms_2000}, though in a previous study we failed to find evidence of a temporal storage effect 
in this community \citep{adler_weak_2009}. Interactions with above- or belowground enemies can regulate populations and, given 
enough species-specificity in the effects of these enemies, decouple the dynamics of different species dynamics \citep{hersh_evaluating_2011,
janzen_herbivores_1970,connell_role_1971}.  More likely, plants may alter the environment, and even resource availability, 
in ways that facilitate their neighbors. For example, shade cast by \textit{A. tripartita} canopies may reduce 
evapotranspiration (e.g. \citealt{Barbier2008}). \textit{A. tripartita} removal might offset this facilitative effect by reducing 
resource uptake, resulting in little net change in water balance following removal. 
Unfortunately, we did not monitor soil moisture at the fine spatial and temporal resolution necessary to test this hypothesis.

>>>>

In spring 2011, in a large exclosure containing six of the historical permanent quadrats, we established an additional 16 quadrats for the removal treatments. New quadrats were selected to be similar to historical plots, but also needed to have enough of the focal species present so that removal experiments could be conducted. We evaluated potential locations for the new quadrats near the original quadrats and then used only those locations with at least 5\% canopy cover of \textit{Artemisia tripartita} and at least 3\% combined basal cover of the three focal perennial grasses. We rejected locations falling on hill slopes, areas with greater than 20\% bare rock, or with over 10\% cover of the woody shrubs \textit{Purshia tridentata} or \textit{Amelanchier utahensis}. We mapped the new quadrats in June, 2011 and then implemented the removal treatments in September, 2011. We randomly assigned 8 of the sixteen new quadrats to the sagebrush removal treatment, and the remaining 8 to the grass removal treatment. We cut sagebrush stems at ground level and applied a 5\% solution of Roundup Pro (Glyphosate, Monsanto) to the cut stems. In the grass removal plots, we painted the same herbicide solution on all perennial grasses. We removed plants 50 cm beyond the quadrat boundaries to minimize edge effects. The sagebrush removal was virtually 100\% effective; grass removals required additional herbicide applications in April and May of each year, with fewer grasses remaining each year. We repeated mapped censuses of all plots each June from 2012-2016.

@@

Our regression models for survival, growth and recruitment follow previous work \citep{adler_coexistence_2010,chu_large_2015}. The current analysis departs from previous work with the inclusion of a categorical removal treatment effect. Although the three vital rate models differ in functional form (see Appendix A for complete model descriptions), all of them have a linear predictor, $\mu$, containing the following terms:
\begin{equation}
\mu_{ij} = \gamma_j+ \varphi_{jg}+  \chi_j  + \sum \limits_{m} \omega_{jm} {W}_{ijm}
\label{eqn:survReg}
\end{equation}
where $i$ refers to an individual genet (for survival and growth) or quadrat (for recruitment) of species $j$, $\gamma$ is a time-dependent intercept (year effect), $\varphi$ is the coefficient for the effect of spatial location (`$g$' represents a group of nearby quadrats located within one pasture or grazing exclosure), $\chi$ is the removal treatment effect, 
$W_{ijm,t}$ is the crowding on species $j$ exerted by neighbors of species $m$ at the individual (survival and growth) or
quadrat (recruitment) scale, and $\omega_{jm}$ is an interaction coefficient which determines the impact of crowding by species $m$ on the 
focal species $j$. For plants in removal plots in 2012, we set crowding for removed neighbor species to zero, rather than using the pre-treatment (2011) data.

@@

The story is different for the fourth species, \textit{P. spicata}. Both individual growth and projected short- and long-term population 
growth were higher in the sagebrush removal treatments than in the control plots, suggesting that our analyses of observational data may 
underestimate the competitive effect of \textit{A. tripartita} on \textit{P. spicata}, and the resulting positive effect of \textit{A. tripartita} 
removal. However, the mechanism by which \textit{A. tripartita} removal promotes \textit{P. spicata} remains unclear. If  \textit{A. tripartita} 
removal is releasing \textit{P. spicata} from competition, we should expect \textit{P. spicata} individuals that were located beneath \textit{A. tripartita} 
shrubs that we removed to respond more than individuals located away from any shrubs. Yet we found no evidence for such a difference. \new{Perhaps 
\textit{A. tripartita's} roots extend far enough beyond its canopy to compete for resources with grasses located in interspaces.}
An alternative explanation involves pre-treatment differences in composition between control plots and  the \textit{A. tripartita} 
removal plots, which had higher cover of \textit{A. tripartita} (see 2011 means in Fig. \ref{fig:CoverTrends}) and higher cover of subdominant 
grasses and shrubs (Fig. \ref{fig:W-by-treatment}). Perhaps these pre-treatment differences reflect subtle differences in 
edaphic conditions which could promote \textit{P. spicata} growth. In summary, while \textit{P. spicata} appears to 
show stronger competitive release than our baseline model predicted, the evidence is not conclusive.

How can we reconcile the insensitivity of grasses to sagebrush removal in our study with evidence from rangeland ecology 
research showing that sagebrush eradication increases grass production? The most likely explanation involves ecological context: our 
study focused on \textit{Artemisia tripartita} in an ungrazed plant community, whereas most research on sagebrush 
and forage production has focused on \textit{Artemisia tridentata} in heavily grazed communities, and often has involved grass seeding 
following sagebrush removal (e.g. \citealt{Cook1963}). \textit{A. tripartita} is smaller in stature than \textit{A. tridentata} 
and may interact differently with perennial grasses. More importantly, cover of sagebrush can increase dramatically on overgrazed sites, 
especially under a spring grazing regime. Competitive release of grasses following sagebrush removal often increases in 
a nonlinear fashion, with little increase in production at sites with low initial sagebrush cover, 
and accelerating responses as initial sagebrush cover passes some threshold \citep{McDaniel2005}, \new{implying a shift in the nature of interspecific interactions, perhaps due to changes in the balance of competition and facilitation or in resource supply.} The cover of 
sagebrush in our ungrazed quadrats may have fallen below that threshold.

The weak interspecific interactions among many of the dominant species in our system implies that these species  
are limited by quite different factors. This low niche overlap could reflect strong resource partitioning \citep{tilman_resource_1982}, 
perhaps combined with variation in the times and locations at which species are taking up those resources. 
Species-specific responses to spatial and temporal variation in non-resource environmental variables can also weaken interspecific 
interactions \citep{chesson_mechanisms_2000}, though in a previous study we failed to find evidence of a temporal storage effect 
in this community \citep{adler_weak_2009}. Interactions with above- or belowground enemies can regulate populations and, given 
enough species-specificity in the effects of these enemies, decouple the dynamics of different species \citep{hersh_evaluating_2011,
janzen_herbivores_1970,connell_role_1971}.  More likely, plants may alter the environment, and even resource availability, 
in ways that facilitate their neighbors. For example, shade cast by \textit{A. tripartita} canopies may reduce 
evapotranspiration (e.g. \citealt{Barbier2008}). \textit{A. tripartita} removal might offset this facilitative effect by reducing 
resource uptake, resulting in little net change in water balance following removal. 
Unfortunately, we did not monitor soil moisture at the fine spatial and temporal resolution necessary to test this hypothesis.
